\Huge Literature Search Made Easy: Mining Online Archives for Research Topic Trends

### 2013-12-13 doc/article_1.tex
Update figures of hLDA and DTM
Updated report with term recounting

  \item Draw an L dimensional topic proportion vector $\theta$ from $Dir(\alpha)$.
  \item For each word $n \in \{ 1, \ldots, N \}$:
  	\begin{enumerate}
  	  	\item Draw $z \in \{ 1, \ldots, L \}$ from {\it Multinomial($\theta$)}.
    		\item Draw $w_n$ from the topic associated with $c_z$.
  	\end{enumerate}
\end{enumerate}
,where the root topic is the topic on level 1, its sub-topics are those on level 2, etc.
			\begin{figure}[!ht]
				\centerline{\includegraphics[scale = 0.8]{hLDA_paper.eps}}
				\caption{Illustration of generative process of hLDA \cite{2}}
				\label{fig: hLDA}
			\end{figure}

@@

\section*{Data Collection}
\subsection*{Data Source}

\subsection*{Feature Extraction}
The collection of abstracts have to be converted to a term frequency matrix as required by all the topic models described above. Instead of generating a term matrix with raw data, we pre-process the data as follows to remove noise:
\begin{enumerate}
  \item Separate an abstract into sentences.
  \item For each sentence:
    	\begin{enumerate}
  	  	\item Remove stop words (e.g. 'the', 'and', 'of').
		\item Remove tokens containing no letters.
		\item Remove common verbs (e.g. 'get', 'take', 'think').
		\item Remove tokens shorter than minimum term length of 3 characters.
		\item Remove latex format tokens.
		\item Split hyphen connected words if both words are longer than the minimum term length.
		\item Convert nouns to sigular form and verbs to infinitive form.
		\item Stem tokens.
  	\end{enumerate}
\end{enumerate}
Then we generate a dictionary containing all terms that appeared in at least 5 documents and a term matrix whose (i, j)-th element n denotes that the j-th word in the dictionary occurred n times in the i-th document.

\subsection*{Feature Transformation}
While doing data pre-processing, we observed that new concepts in science and technology are usually expressed in the form of phrases, such as "computer science", "machine learining" and "topic modeling", which indicates a topic might be better characterized by its "key phrases" instead of keywords. Furthermore, phrases also offer better interpretability of results, e.g. a topic with keywords {"dimension", "principal", "analysis"} may greatly perplex us, while one with key phrases {"principal component analysis", "dimension reduction"} is quite informative.\newline
We generated phrases before the second step in the previous section by:
\begin{enumerate}
  \item Separate words into blocks by punctuation marks and stop words.
  \item Within each block, generate phrases made up of consecutive words while restricting the maximum length of phrases to 5 words since usually phrases longer than 5 words are represented by their acronyms.
\end{enumerate}

@@

\subsection*{Results}
\subsubsection*{Dynamic Topic Model}
In our DTM experiment, the number of topics is set to 10 a priori. Results are summarized in Figure ~\ref{fig: DTM_table_unweight} and Figure ~\ref{fig: PhaseTrans_wp}. In Figure ~\ref{fig: DTM_table_unweight} we demonstrate 5 out of the 10 topics for purpose of figure size control, the 3 tables respectively are topics and keywords generated from only words, words plus phrases and only phrases. The classification with phrases clearly gives much better interpretability. Note that the first two tables look quite similar, in fact, in the second table, even though the corpus contains both words and phrases, we can hardly see any phrases among the important keywords. This is quite intuitive since in the corpus, there are 486712 words, 4645 of them unique and 112448 phrases with 11250 of them unique. So on average, a word appears ~100 times in the corpus while a phrase appears only ~10 times, resulting in a term matrix dominated by words. This is what has motivated our weighting scheme. \begin{figure}[!ht]
  \includegraphics[scale = 0.45]{dtm10_unweight.eps}
  \caption{DTM Topics and Keywords}
  \label{fig: DTM_table_unweight}
\end{figure}
In Figure ~\ref{fig: DTM_table_unweight}, we present the plot of one specific topic "phase transition", which is typical of all the plots.
\begin{figure}[!ht]
  \includegraphics[scale = 0.365]{PhaseTrans(w_p).eps}
  \caption{Keywords Trend Plot of Topic "Phase Transition"}
  \label{fig: PhaseTrans_wp}
\end{figure}[!ht]
From Figure ~\ref{fig: DTM_table_unweight}, despite the frequencies of keywords do rise and fall, there is hardly any obvious "trend". This is not unexpected since it's natural that a paper on "phase transition" should always involve keywords such as "phase transition" and "phase diagram" regardless of when it is published.\newline
One very informative plot we can make is the evolution of topic popularity. In Figure ~\ref{fig: TopPop_p}, we show a plot of frequencies for 4 topics over time.
\begin{figure}[!ht]
  \includegraphics[scale = 0.365]{TopPop_p.eps}
  \caption{Plot of Frequencies of Different Topics over Time}
  \label{fig: TopPop_p}
\end{figure}
This plot has some really interesting implications: 
\begin{inparaenum}[\itshape a\upshape)]
\item "Superconductivity" and "Ultracold Atoms" are clearly negatively correlated, which indicates it is possible that for some reason (e.g. similarities in research techniques) research works in these two fields are almost done by the same group of researchers. The negative correlation is created by the shift of their interest and time allocation between the two fields. So for a Phd student whose thesis project is in superconductivity, it may be wise to choose a backup side project in "Ultracold Atoms", in the case her thesis project fails, she can turn to the side project and produce some results quickly without having to acquire a new set of skills.
\item In certain field, it is easy to publish, but difficult to discover something of importance, while it's the opposite case in some other fields: "Quantum Hall Effect", although being quite famous even outside the field of physics has actually very few publications, which is probably due to the requirement of ultra high magnetic field in quantum Hall experiments. On the other hand, there are surprisingly large number of papers on "Computations and Simulations" despite the small number of researchers in this field (e.g. one out of about 40 physics faculty members at Stanford). This implies that if one's priority is to have as much publication as possible, then he should choose to do simulations.
\end{inparaenum}

\subsubsection*{Hierarchical Latent Dirichlet Allocation}
 We produced three depth 4 trees generated words alone, words plus phrases and phrases alone, where the depth has to be chosen a priori. Branches with less than 30 keywords are pruned off. In Figure 7, we show the ** tree up to the fourth level, while the other two to the third level, because there are already too many branches on the third level. tree generated by words only, we only show the trees only up to the third level  fourth level to limit size of the plot. We also only display the first three branches from each node to limit the size of the tree, there are usually 5 trees from each node.

@@

\section*{Discussion}

\section*{Acknowlegement}
We would like to thank the TAs for their guidance and advice for this project. We also want to thank Prof. Andrew Ng for introducing us to the exciting field of machine learning.  
%----------------------------------------------------------------------------------------
%	REFERENCE LIST
%----------------------------------------------------------------------------------------

>>>>

  \item Draw an L dimensional topic proportion vector $\theta$ from $Dir(\alpha)$.
  \item For each word $n \in \{ 1, \ldots, N \}$:
  	\begin{enumerate}
  	  	\item Draw $z \in \{ 1, \ldots, L \}$ from {\it Multinomial($\theta$)}.
    		\item Draw $w_n$ from the topic associated with $c_z$.
  	\end{enumerate}
\end{enumerate}
where the root topic is the topic on level 1, its sub-topics are those on level 2, etc.
			\begin{figure}[!ht]
				\centerline{\includegraphics[scale = 0.8]{hLDA_paper.eps}}
				\caption{Illustration of generative process of hLDA \cite{2}}
				\label{fig: hLDA}
			\end{figure}

@@

\section*{Feature Extraction}
\subsection*{Data Cleanse}
The collection of documents have to be converted to a term frequency matrix as required by all the topic models described above. Instead of generating a term matrix with raw data, we pre-process the data as follows to remove noise:
\begin{enumerate}
  \item Separate an abstract into sentences.
  \item For each sentence:
    	\begin{enumerate}
  	  	\item Remove stop words (e.g. 'the', 'and', 'of').
		\item Remove tokens containing no letters.
		\item Remove common verbs (e.g. 'get', 'take', 'think').
		\item Remove tokens shorter than minimum term length of 3 characters.
		\item Remove latex format tokens.
		\item Split hyphen connected words if both words are longer than the minimum term length.
		\item Convert nouns to sigular form and verbs to infinitive form.
		\item Stem tokens.
  	\end{enumerate}
\end{enumerate}
Then we generate a dictionary containing all terms that appeared in at least 5 documents and a term matrix whose ($i$, $j$)-th element n denotes that the $j$-th word in the dictionary occurred n times in the $i$-th document.

\subsection*{Phrase Generation}
While doing data pre-processing, we observed that new concepts in science and technology are usually expressed in the form of phrases, such as "computer science", "machine learning" and "topic modeling", which indicates a topic might be better characterized by its "key phrases" instead of keywords. Furthermore, phrases also offer better interpretability of results, e.g. a topic with keywords {"dimension", "principal", "analysis"} may greatly perplex us, while one with key phrases {"principal component analysis", "dimension reduction"} is quite informative.\newline
We generated phrases before the second step in the previous section by:
\begin{enumerate}
  \item Separate words into blocks by punctuation marks and stop words.
  \item Within each block, generate phrases made up of consecutive words while restricting the maximum length of phrases to 5 words since usually phrases longer than 5 words are represented by their acronyms.
\end{enumerate}
Using the phrase generation method described above can lead to duplicate counts in the result. For example, phrase "Stanford University" will generate "Stanford":1, "University":1 and "Stanford University":1, while these two terms should only be counted as a phrase once. To eliminate this inaccuracy, we implemented a two-phase algorithm. The first phase is the same as the approach described above: extract terms and enumerate phrases in each document, then filter the terms and phrases whose occurrences are below the minimum threshold. In the second phase, documents are scanned through again. For each term in a document, we check whether it can be combined with its previous or next term to generate a phrase appearing in the filtered phrase dictionary. If not, this term is counted as one individual occurance; otherwise this term is skipped. This method basically recounts all terms and checks if their appearance contributing to the generation of a phrase. Only those terms not as part of a phrase is counted.

@@

\subsection*{Results}
\subsubsection*{Trends in Research Topics}
In our DTM experiment, the number of topics is set to 10 a priori. Results are summarized in Figure ~\ref{fig: DTM_table_unweight} and Figure ~\ref{fig: PhaseTrans_wp}. In Figure ~\ref{fig: DTM_table_unweight} we demonstrate 5 out of the 10 topics for purpose of figure size control, the 3 tables respectively are topics and keywords generated from only words, words plus phrases and only phrases. The classification with phrases clearly gives much better interpretability. Note that the first two tables look quite similar, in fact, in the second table, even though the corpus contains both words and phrases, we can hardly see any phrases among the important keywords. This is quite intuitive since in the corpus, there are 486712 words, 4645 of them unique and 112448 phrases with 11250 of them unique. So on average, a word appears ~100 times in the corpus while a phrase appears only ~10 times, resulting in a term matrix dominated by words. This is what has motivated our weighting scheme. \begin{figure}[!ht]
  \includegraphics[scale = 0.45]{dtm10_unweight.eps}
  \caption{DTM Topics and Keywords}
  \label{fig: DTM_table_unweight}
\end{figure}
In Figure ~\ref{fig: DTM_table_unweight}, we present the plot of one specific topic "phase transition", which is typical of all the plots.
\begin{figure}[!ht]
  \includegraphics[scale = 0.365]{PhaseTrans(w_p).eps}
  \caption{Keywords Trend Plot of Topic "Phase Transition"}
  \label{fig: PhaseTrans_wp}
\end{figure}[!ht]
From Figure ~\ref{fig: DTM_table_unweight}, despite the frequencies of keywords do rise and fall, there is hardly any obvious "trend". This is not unexpected since it's natural that a paper on "phase transition" should always involve keywords such as "phase transition" and "phase diagram" regardless of when it is published.\newline
One very informative plot we can make is the evolution of topic popularity. In Figure ~\ref{fig: TopPop_p}, we show a plot of frequencies for 4 topics over time.
\begin{figure}[!ht]
  \includegraphics[scale = 0.365]{TopPop_p.eps}
  \caption{Plot of Frequencies of Different Topics over Time}
  \label{fig: TopPop_p}
\end{figure}
This plot has some really interesting implications: 
\begin{inparaenum}[\itshape a\upshape)]
\item "Superconductivity" and "Ultracold Atoms" are clearly negatively correlated, which indicates it is possible that for some reason (e.g. similarities in research techniques) research works in these two fields are almost done by the same group of researchers. The negative correlation is created by the shift of their interest and time allocation between the two fields. So for a Phd student whose thesis project is in superconductivity, it may be wise to choose a backup side project in "Ultracold Atoms", in the case her thesis project fails, she can turn to the side project and produce some results quickly without having to acquire a new set of skills.
\item In certain field, it is easy to publish, but difficult to discover something of importance, while it's the opposite case in some other fields: "Quantum Hall Effect", although being quite famous even outside the field of physics has actually very few publications, which is probably due to the requirement of ultra high magnetic field in quantum Hall experiments. On the other hand, there are surprisingly large number of papers on "Computations and Simulations" despite the small number of researchers in this field (e.g. one out of about 40 physics faculty members at Stanford). This implies that if one's priority is to have as much publication as possible, then he should choose to do simulations.
\end{inparaenum}

\subsubsection*{Topic Hierarchy and Key Phrases}
 We produced three depth 4 trees generated words alone, words plus phrases and phrases alone, where the depth has to be chosen a priori. Branches with less than 30 keywords are pruned off. In Figure 7, we show the ** tree up to the fourth level, while the other two to the third level, because there are already too many branches on the third level. tree generated by words only, we only show the trees only up to the third level  fourth level to limit size of the plot. We also only display the first three branches from each node to limit the size of the tree, there are usually 5 trees from each node.

@@

\section*{Conclusion}

%----------------------------------------------------------------------------------------
%	REFERENCE LIST
%----------------------------------------------------------------------------------------
